---
title: "Reintroduction of Wolves in Yellowstone"
author: "Marcus Varni, Ale Vidal Meza, and Emily Sorenson"
date: "2025-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Johnson et al. (2015) revised the Lotka-Volterra predator-prey model:
$$
Elk\:\:population\:\:over\:\:time
\newline
\begin{aligned} \frac{dE}{dt} &= k_{p1} E - p_{ec} E C - p_{ew} E W\end{aligned}\\\
\newline
Coyote\:\:population\:\:over\:\:time
\newline
\begin{aligned} \frac{dC}{dt}&= -k_{d1} C + k_{p2} E C - k_{d2} C W\end{aligned}\\\
\newline
Wolf\:\:population\:\:over\:\:time
\newline
\begin{aligned} \frac{dW}{dt} &= -k_{d3} W + k_{p3}EW\end{aligned}
$$

```{r}
# Set variables based on conference paper
E0 <- 1.67 * (10^(4)) # Define initial elk population
C0 <- 80 # Define initial coyote population
W0 <- 21 # Define initial wolf population

# Set parameters based on conference paper
kp1 <- 0.467  # Define elk growth rate
pec <- 7.88 * (10^(-3)) # Define predation rate of coyotes on elk
pew <- 7.16 * (10^(-3)) # Define predation rate of wolves on elk
kd1 <- 0.4251 # Define natural decay of coyotes given lack of food
kp2 <- 2.56 * (10^(-4)) # Define proliferation rate of coyotes
kd2 <- 4.96 * (10^(-7)) # Define hostile incidents between wolves and coyotes since reintroduction
kd3 <- 2.81 * (10^(-7)) # Define natural decay rate of wolves 
kp3 <- 0.015 # Define proliferation rate of wolves given sustenance
```

```{r}
# Check inputs
kp1 # View elk growth rate
pec # View predation rate of coyotes on elk
pew # View predation rate of wolves on elk
kd1 # View natural decay of coyotes given lack of food
kp2 # View proliferation rate of coyotes
kd2 # View hostile incidents between wolves and coyotes since reintroduction
kd3 # View natural decay rate of wolves 
kp3 # View proliferation rate of wolves given sustenance
```

```{r}
# Create time vector of length 100000
tset <- seq(from = 0, to = 35, length.out = 100000)

E.simu <- NaN*tset # Create empty holding vector
E.simu[1] <- E0 # Store initial population value in vector

C.simu <- NaN*tset # Create empty holding vector
C.simu[1] <- C0 # Store initial population value in vector

W.simu <- NaN*tset # Create empty holding vector
W.simu[1] <- W0 # Store initial population value in vector
```

```{r}
# Initiate for loop, start in second slot due to initial condition
for(i in 2:length(tset)){
  dt <- tset[i] - tset[i-1] # Find change in time
  
  # Assign temporary population variables
  E <- E.simu[i-1] # Extract number of elk at previous timestep
  C <- C.simu[i-1] # Extract number of coyote at previous timestep
  W <- W.simu[i-1] # Extract number of wolves at previous timestep
  
  # Calculate population rate of change
  dE <- ((kp1*E) + (-pec*E*C) + (-pew*E*W)) * dt # Find elk population over time
  dC <- ((-kd1*C) + (kp2*E*C) + (-kd2*C*W)) * dt # Find coyote population over time
  dW <- ((-kd3*W) + (kp3*E*W)) * dt # Find wolf population over time
  
  # Calculate population change
  E.simu[i] <- E + dE # Current elk population at timestep
  C.simu[i] <- C + dC # Current coyote population at timestep
  W.simu[i] <- W + dW # Current wolf population at timestep
}
```

```{r}
# Check start/end values of the simulations
head(E.simu)
tail(E.simu)

head(C.simu)
tail(C.simu)

head(W.simu)
tail(W.simu)
```

```{r}
# Set variable colors for plots
Ecol <- "#219ebc"
Ccol <- "#d62828"
Wcol <- "#6a994e"
```

```{r}
# Plot N as a function of time
plot(x = tset, # Add time vector to x axis
     y = E.simu, # Add elk population size to y axis
     type = "l", # Set type to "Lines"
     col = Ecol, # Specify line color
     las = 1, # Change orientation of tick marks to be horizontal
     lwd = 2, # Set line width to 2
     xlab = "Time", # Set x axis label
     ylab = "Number of Individuals") # Set y axis label,
lines(x = tset, # Add time vector to x axis
      y = C.simu, # Add coyote population size to y axis
      col = Ccol, # Specify line color
      lwd = 2) # Set line width to 2
lines(x = tset, # Add time vector to x axis
      y = W.simu, # Add wolf population size to y axis
      col = Wcol, # Specify line color
      lwd = 2) # Set line width to 2
abline(h = 0) # Add horizontal line at x = 0
legend(x = max(tset)*0.1, y = 0.8*E0, # Set legend position
       lwd = 2, # Set line width to 2
       legend = c("Elk pop.", "Coyote pop.", "Wolf pop."), # Set item labels
       col = c(Ecol, Ccol, Wcol), # Set item colors
       horiz = TRUE) # Set legend vertically
```

```{r}
# ylim at 100
# Setting variable colors for plots
plot(x = tset, y = E.simu,
     type = "l", 
     col = Ecol, 
     las = 1, 
     lwd = 2, 
     ylim = c(0,100),
     xlab = "Time", ylab = "Number of Individuals")

lines(x = tset, y = C.simu, col = Ccol, lwd = 2)

lines(x = tset, y = W.simu, col = Wcol, lwd = 2)

abline(h=0)

legend(x = max(tset)*0.1, y = 0.8*E0, 
       lwd = 2, 
       legend = c("Elk pop.", "Coyote pop.", "Wolf pop."),
       col = c(Ecol, Ccol, Wcol), 
       horiz = TRUE)
```

```{r}
# xlim at 1
# Setting variable colors for plots
plot(x = tset, y = E.simu,
     type = "l", 
     col = Ecol, 
     las = 1, 
     lwd = 2, 
     xlim = c(0,1),
     xlab = "Time", ylab = "Number of Individuals")

lines(x = tset, y = C.simu, col = Ccol, lwd = 2)

lines(x = tset, y = W.simu, col = Wcol, lwd = 2)

abline(h=0)

legend(x = max(tset)*0.1, y = 0.8*E0, 
       lwd = 2, 
       legend = c("Elk pop.", "Coyote pop.", "Wolf pop."),
       col = c(Ecol, Ccol, Wcol), 
       horiz = TRUE)
```

Based on simulated results published by Johnson et al. (2015), we expected to see a rise in the wolf population from 1995-2006 and a plateau around 2007-2030; an exponential decline in the coyote population from 1995 to 2003 and onwards; and a decline in the elk population over time from 1995 to 2008 and onwards. However, our plotted time series based on the initial population and parameter values published led to signficantly different results.