---
title: "Reintroduction of Wolves in Yellowstone"
author: "Marcus Varni, Ale Vidal Meza, and Emily Sorenson"
date: "2025-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Johnson et al. (2015) revised the Lotka-Volterra predator-prey model:
$$
Elk\:\:population\:\:over\:\:time
\newline
\begin{aligned} \frac{dE}{dt} &= k_{p1} E - p_{ec} E C - p_{ew} E W\end{aligned}\\\
\newline
Coyote\:\:population\:\:over\:\:time
\newline
\begin{aligned} \frac{dC}{dt}&= -k_{d1} C + k_{p2} E C - k_{d2} C W\end{aligned}\\\
\newline
Wolf\:\:population\:\:over\:\:time
\newline
\begin{aligned} \frac{dW}{dt} &= -k_{d3} W + k_{p3}EW\end{aligned}
$$
```{r}
# Set variable colors for plots
 Ecol <- "#219ebc"
 Ccol <- "#d62828"
 Wcol <- "#6a994e"
```

```{r}
# Set variables based on conference paper
E0 <- 500#1.67 * (10^(4)) # Define initial elk population
C0 <- 200 # Define initial coyote population
W0 <- 200 # Define initial wolf population

# Set parameters based on conference paper
kp1 <- 0.467  # Define elk growth rate
pec <- 0.001#0.0788  # Define predation rate of coyotes on elk
pew <- pec#0.001  # Define predation rate of wolves on elk
kd1 <- 0.04251 # Define natural decay of coyotes given lack of food
kp2 <- 0.1*pec  # Define proliferation rate of coyotes
kd2 <- 0.0005#4.96  # Define hostile incidents between wolves and coyotes since reintroduction
kd3 <- kd1#0.0281  # Define natural decay rate of wolves 
kp3 <- 0.1*pew # Define proliferation rate of wolves given sustenance


# Create time vector of length 100000
tset <- seq(from = 0, to = 200, by = .001)

E.simu <- NaN*tset # Create empty holding vector
E.simu[1] <- E0 # Store initial population value in vector

C.simu <- NaN*tset # Create empty holding vector
C.simu[1] <- C0 # Store initial population value in vector

W.simu <- NaN*tset # Create empty holding vector
W.simu[1] <- W0 # Store initial population value in vector

# Initiate for loop, start in second slot due to initial condition
for(i in 2:length(tset)){
  dt <- tset[i] - tset[i-1] # Find change in time
  
  # Assign temporary population variables
  E <- E.simu[i-1] # Extract number of elk at previous timestep
  C <- C.simu[i-1] # Extract number of coyote at previous timestep
  W <- W.simu[i-1] # Extract number of wolves at previous timestep
  
  # Calculate population rate of change
  dE <- ((kp1*E) + (-pec*E*C) + (-pew*E*W)) * dt # Find elk population over time
  dC <- ((-kd1*C) + (kp2*E*C) + (-kd2*C*W)) * dt # Find coyote population over time
  dW <- ((-kd3*W) + (kp3*E*W)) * dt # Find wolf population over time
  
  # Calculate population change
  E.simu[i] <- E + dE # Current elk population at timestep
  C.simu[i] <- C + dC # Current coyote population at timestep
  W.simu[i] <- W + dW # Current wolf population at timestep
}

# Plot N as a function of time
plot(x = tset, # Add time vector to x axis
     y = E.simu, # Add elk population size to y axis
     type = "l", # Set type to "Lines"
     col = Ecol, # Specify line color
     las = 1, # Change orientation of tick marks to be horizontal
     lwd = 2, # Set line width to 2
     ylim = c(min(c(E.simu,C.simu,W.simu),na.rm=T),max(c(E.simu,C.simu,W.simu),na.rm = T)),
     xlab = "Time", # Set x axis label
     ylab = "Number of Individuals") # Set y axis label,
lines(x = tset, # Add time vector to x axis
      y = C.simu, # Add coyote population size to y axis
      col = Ccol, # Specify line color
      lwd = 2) # Set line width to 2
lines(x = tset, # Add time vector to x axis
      y = W.simu, # Add wolf population size to y axis
      col = Wcol, # Specify line color
      lwd = 2) # Set line width to 2
abline(h = 0) # Add horizontal line at x = 0
legend(x = 'topleft',inset=c(0,0), # Set legend position
       lwd = 2, # Set line width to 2
       legend = c("Elk pop.", "Coyote pop.", "Wolf pop."), # Set item labels
       col = c(Ecol, Ccol, Wcol), # Set item colors
       horiz = TRUE) # Set legend vertically
```

```{r}
# Check start/end values of the simulations
head(E.simu)
tail(E.simu)

head(C.simu)
tail(C.simu)

head(W.simu)
tail(W.simu)
```





```{r}
##Bifurcation diagram

## a. Selecting the x-axis for our bifurcation diagram
kd2_set <- seq(from = 0, to = 100, by = 50) 

## b. Calculating the equilibrium points for each value of Y


## Storage variables for our two state variables of interest. 

Estarset <- NaN*kd2_set # holding vector for equilibrium prey population size corresponding to each value of Y in Yset 
Cstarset <- NaN*kd2_set # holding vector for equilibrium prey population size corresponding to each value of Y in Yset 
Wstarset <- NaN*kd2_set # holding vector for equilibrium prey population size corresponding to each value of Y in Yset 
### For loop

for(j in 1:length(kd2_set)){ # for each element j from 1 to the length of Yset
  # assign the value of Y
	kd2 <- kd2_set[j] # set Y equal to the jth element of Yset
	
	# now simulate the model with this value of Y:
	# create holding vectors for predator and prey populations and fill in initial conditions
	E.simu <- NaN*tset # holding vector for the prey population 
	E.simu[1] <- 1 # fill in initial prey size
	C.simu <- NaN*tset # holding vector for the predator population
	C.simu[1] <- 1 # fill in initial predator size
	W.simu <- NaN*tset # holding vector for the predator population
	W.simu[1] <- 1 # fill in initial predator size
	
	for(i in 2:length(tset)){ # for each element i from 2 to length of tset
	  # calculating change in time
		dt <- tset[i]-tset[i-1]
		
		# storing temporary variables
		E <- E.simu[i-1] # call the prey population size at the previous timepoint X
		C <- C.simu[i-1] # call the predator population size at the previous timepoint X
		W <- W.simu[i-1]
		
		 # Calculate population rate of change
  dE <- ((kp1*E) + (-pec*E*C) + (-pew*E*W)) * dt # Find elk population over time
  dC <- ((-kd1*C) + (kp2*E*C) + (-kd2*C*W)) * dt # Find coyote population over time
  dW <- ((-kd3*W) + (kp3*E*W)) * dt # Find wolf population over time
		
		 # Calculate population change
  E.simu[i] <- E + dE # Current elk population at timestep
  C.simu[i] <- C + dC # Current coyote population at timestep
  W.simu[i] <- W + dW # Current wolf population at timestep
		
	} # end of model simulation
	
	# storing last population size (equilibrium population size) in holding the "starset" holding vectors
	Estarset[j] <- E.simu[length(tset)] # prey population size at the end of the simulation with Y = Yset[j]
	Cstarset[j] <- C.simu[length(tset)] # predator population size at the end of the simulation with Y = Yset[j]
	Wstarset[j] <- W.simu[length(tset)] # predator population size at the end of the simulation with Y = Yset[j]
	
}# end of our outer for loop that iterates over each element in kd2_set

## c. Examining the results

## We can now make bifurcation diagrams for both X and P.

# plot stable equilibria of predator (Pstarset) against density of alternate prey (Yset)
plot(x = kd2_set, y = Estarset, # x = density of alternate prey, y = equilibrium predator population size
     type = 'l', lwd = 2, col = Ecol, las = 1, # plot aesthetics
     xlab = 'Hostile Incidence Between Wolves and Coyotes, kd2', ylab = 'Stable equilibria of elk, E*') # axis labels

# plot stable equilibria of prey (Xstarset) against density of alternate prey (Yset)
plot(x = kd2_set, y = Cstarset, # x = density of alternate prey, y = equilibrium prey population size
     type = 'l', lwd = 2, col = Ccol, las = 1, # plot aesthetics
     xlab = 'Hostile Incidence Between Wolves and Coyotes, kd2', ylab = 'Stable equilibria of coyotes, C*')# axis labels

# plot stable equilibria of prey (Xstarset) against density of alternate prey (Yset)
plot(x = kd2_set, y = Wstarset, # x = density of alternate prey, y = equilibrium prey population size
     type = 'l', lwd = 2, col = Wcol, las = 1, # plot aesthetics
     xlab = 'Hostile Incidence Between Wolves and Coyotes, kd2', ylab = 'Stable equilibria of Wolves, W*')# axis labels
```

```{r}
# xlim at 1
head(Estarset)
tail(Estarset)
```

Based on simulated results published by Johnson et al. (2015), we expected to see a rise in the wolf population from 1995-2006 and a plateau around 2007-2030; an exponential decline in the coyote population from 1995 to 2003 and onwards; and a decline in the elk population over time from 1995 to 2008 and onwards. However, our plotted time series based on the initial population and parameter values published led to signficantly different results.