---
title: "The Reintroduction of Wolves in Yellowstone: A Multi-Predator Model"
author: "Marcus Varni, Alessandra Vidal Meza, and Emily Sorenson"
date: "2025-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(tidyverse)
```

## Background

Trophic cascades refer to the indirect effects of the addition or removal of a top predator in an ecosystem or food web. In 1926, grey wolves (Canis lupus) were eradicated from Yellowstone National Park (YNP) after decades of livestock conflict with westward settlers, leading to an ecosystem collapse (History of Wolf Management in Yellowstone National Park, 2025). The reintroduction of wolves into YNP in 1995 is an emblematic case of successful restoration, which resulted in a trophic cascade and reinstated top-down control of herbivory in the park. Without a primary predator, the Rocky Mountain elk (Cervus canadensis) population had exploded into an almost exponential growth and exerted herbivory pressure on woody plants like willows and aspens from overgrazing (Ripple & Beschta, 2012). Similarly, the eradication of wolves in the park was linked to the expansion and increased abundance of coyotes (Canis latrans) (Ripple et al., 2013). Coyotes are a mesopredator and scavenger species that often feeds on wolf-killed ungulate carcasses. Coyotes are typically suppressed by wolves and occur at low densities across wolf-dominated landscapes (Merkle et al., 2009; Ripple et al., 2013).

Several mathematical biological models have been constructed in an attempt to better understand trophic cascades in large and complex ecosystems like YNP and inform future management. In a conference paper published by Johnson et al. (2015), a mathematical biology model based on the Lotka-Volterra predation model is proposed to describe multi-predator dynamics and capture the interactions between elk, wolves, and coyotes in YNP. The authors novely perform a Markov Chain Monte Carlo (MCMC) analysis and apply a Metropolis-Hastings algorithm to calibrate the model parameters based on field observations. We present and discuss this model further here. 

## Model Set Up

Johnson et al. (2015) proposed the following predator-prey model:

**Elk population over time**:

$$
\begin{aligned} \frac{dE}{dt} &= k_{p1} E - p_{ec} E C - p_{ew} E W\end{aligned}\\\
$$
**Coyote population over time**: 
$$
\begin{aligned} \frac{dC}{dt}&= -k_{d1} C + k_{p2} E C - k_{d2} C W\end{aligned}\\\
$$
**Wolf population over time**:
$$
\begin{aligned} \frac{dW}{dt} &= -k_{d3} W + k_{p3}EW\end{aligned}
$$

```{r echo=FALSE}
table <- tibble(
  Variables.And.Parameters = c("W", "E", "C", "kp1", "pec", "pew", "kd1", "kp2", "kd2", "kd3", "kp3"), 
  Description = c("Wolf population", "Elk population", "Coyote population", "Growth rate of elk",
                  "Predation rate of coyotes on elk", "Predation rate of wolves on elk",
                  "Natural decay rate of coyotes without prey", "Proliferation rate of coyotes",
                  "Hostile incidents between wolves and coyotes since reintroduction",
                  "Natural decay rate of wolves without prey", "Proliferation rate of wolves"), 
  Units = c("Number of wolves", "Number of elk", "Number of coyotes", "Elk per year", 
            "Fraction of elk per coyote per year", "Fraction of elk per wolf per year",
            "Fraction of coyotes per year", "Coyotes per year", "Incidents per year",
            "Fraction of wolves per year", "Wolves per year"),
  Johnson.et.al.Value = c("21", "1.67x10^4", "80", "0.467", "7.88 × 10^−3", "7.16 × 10^−3", "0.4251", 
                    "2.56 × 10^−4", "4.96 × 10^−7", "2.81 ×10^−7", "0.015"),
  Modified.Value = c("200", "500", "200", NA, "0.001", "0.001", NA, "0.0001", "0.0005", "0.04251", "0.0001"))

table %>%
  kbl(col.names = gsub("[.]", " ", names(table)),
      caption = "Table 1. Summary of variables and parameters with modified values.") %>%
  kable_styling()
```

```{r}
# Set variables based on conference paper
E0 <- 500 # Define initial elk population
C0 <- 200 # Define initial coyote population
W0 <- 200 # Define initial wolf population

# Set parameters based on conference paper
kp1 <- 0.467  # Define elk growth rate
pec <- 0.001 # Define predation rate of coyotes on elk
pew <- pec # Define predation rate of wolves on elk
kd1 <- 0.04251 # Define natural decay of coyotes given lack of food
kp2 <- 0.1*pec # Define proliferation rate of coyotes
kd2 <- 0.0005 # Define hostile incidents between wolves and coyotes since reintroduction
kd3 <- kd1 # Define natural decay rate of wolves 
kp3 <- 0.1*pew # Define proliferation rate of wolves given sustenance
```

```{r}
# Check inputs
kp1 # View elk growth rate
pec # View predation rate of coyotes on elk
pew # View predation rate of wolves on elk
kd1 # View natural decay of coyotes given lack of food
kp2 # View proliferation rate of coyotes
kd2 # View hostile incidents between wolves and coyotes since reintroduction
kd3 # View natural decay rate of wolves 
kp3 # View proliferation rate of wolves given sustenance
```

```{r}
# Set variable colors for plots
Ecol <- "#219ebc" 
Ccol <- "#d62828"
Wcol <- "#6a994e"
```

## Time Series Simulation

```{r}
# Create time vector from 0 to 200 with timesteps at .001 point
tset <- seq(from = 0, to = 200, by = .001)

E.simu <- NaN*tset # Create empty holding vector
E.simu[1] <- E0 # Store initial population value in vector

C.simu <- NaN*tset # Create empty holding vector
C.simu[1] <- C0 # Store initial population value in vector

W.simu <- NaN*tset # Create empty holding vector
W.simu[1] <- W0 # Store initial population value in vector
```

```{r}
# Initiate for loop, start in second slot due to initial condition
for(i in 2:length(tset)){
  dt <- tset[i] - tset[i-1] # Find change in time
  
  # Assign temporary population variables
  E <- E.simu[i-1] # Extract number of elk at previous timestep
  C <- C.simu[i-1] # Extract number of coyote at previous timestep
  W <- W.simu[i-1] # Extract number of wolves at previous timestep
  
  # Calculate change in population size
	dE <- ((kp1*E) + (-pec*E*C) + (-pew*E*W)) * dt # Change in elk population size over dt
	dC <- ((-kd1*C) + (kp2*E*C) + (-kd2*C*W)) * dt # Change in coyotes population size over dt
	dW <- ((-kd3*W) + (kp3*E*W)) * dt # Change in wolves population size over dt
  
  # Calculate population change
  E.simu[i] <- E + dE # Current elk population at timestep
  C.simu[i] <- C + dC # Current coyote population at timestep
  W.simu[i] <- W + dW # Current wolf population at timestep
}
```

```{r}
# Check start and end values of simulated population values
head(E.simu) # View first 5 values of number of elk
tail(E.simu) # View last 5 values of number of elk

head(C.simu) # View first 5 values of number of coyote
tail(C.simu) # View last 5 values of number of elk

head(W.simu) # View first 5 values of number of wolves
tail(W.simu) # View last 5 values of number of wolves
```

```{r}
# Plot N as a function of time
plot(x = tset, # Add time vector to x axis
     y = E.simu, # Add elk population size to y axis
     type = "l", # Set type to "Lines"
     col = Ecol, # Specify line color
     las = 1, # Change orientation of tick marks to be horizontal
     lwd = 2, # Set line width to 2
     ylim = c(min(c(E.simu, C.simu, W.simu), na.rm = T),
             max(c(E.simu, C.simu, W.simu), na.rm = T)), # Set y axis limit
     main = "Multi-Predator Interactions in Yellowstone National Park", # Set main title
     sub = "Time Series Simulation Adapted from Johnson et al. (2015) for Wolf Reintroduction in 1995",
     xlab = "Time", # Set x axis label
     ylab = "Number of Individuals") # Set y axis label,
lines(x = tset, # Add time vector to x axis
      y = C.simu, # Add coyote population size to y axis
      col = Ccol, # Specify line color
      lwd = 2) # Set line width to 2
lines(x = tset, # Add time vector to x axis
      y = W.simu, # Add wolf population size to y axis
      col = Wcol, # Specify line color
      lwd = 2) # Set line width to 2
abline(h = 0) # Add horizontal line at x = 0
legend(x = "topright", # Set legend position
       lwd = 2, # Set line width to 2
       legend = c("Elk pop.", "Coyote pop.", "Wolf pop."), # Set item labels
       col = c(Ecol, Ccol, Wcol), # Set item colors
       horiz = TRUE) # Set legend vertically
```
```{r}
max_E.simu <- max(E.simu[(length(E.simu)/2):(length(E.simu))])
max_E.simu

min_E.simu <- min(E.simu[(length(E.simu)/2):(length(E.simu))])
min_E.simu

max_W.simu <- max(W.simu[(length(W.simu)/2):(length(W.simu))])
max_W.simu

min_W.simu <- min(W.simu[(length(W.simu)/2):(length(W.simu))])
min_W.simu
```

## Hopf Bifurcation Diagrams 

### Predation Rate of Wolves on Elk

```{r}
# Create x axis vector of range of `pew` values to bifurcate 
pew_set <- seq(from = 0, to = 30, by = 0.5) 

# Create empty holding vectors for state variables to store min/max population sizes 
min_E.starset <- NaN*pew_set # NaN vector of vector length pew_set to store min elk population size
max_E.starset <- NaN*pew_set # NaN vector of vector length pew_set to store max elk population size
min_W.starset <- NaN*pew_set # NaN vector of vector length pew_set to store min wolf population size
max_W.starset <- NaN*pew_set # NaN vector of vector length pew_set to store max wolf population size
```

```{r}
## Calculate equilibrium values for each value of pew in `pew_set` ##
for(j in 1:length(pew_set)){ # Iterate over number of items given by length of pew_set; start outer for loop

	# pew <- pew_set[j] # Assign jth element of pew_set to pew
	pew <- pew_set[j]
	
	## Simulate model for jth element of pew_set ##
	
	# Set initial condition
	E.simu <- NaN*tset # Create NaN holding vector for elk population
	E.simu[1] <- 1 # Set initial elk population size 
	C.simu <- NaN*tset # Create NaN holding vector for coyotes population
	C.simu[1] <- 1 # Set initial coyote population size 
	W.simu <- NaN*tset # Create NaN holding vector for wolves population
	W.simu[1] <- 1 # Set initial wolves population size 
	
	# Initiate for loop, start in second slot due to initial condition; start inner for loop
	for(i in 2:length(tset)){
	  
	  dt <- tset[i]-tset[i-1] # Change in time; timestep
		
		# Create temporary variable from previous timestep
		E <- E.simu[i-1] # Set elk population size at previous timestep
		C <- C.simu[i-1] # Set coyotes population size at previous timestep
		W <- W.simu[i-1] # Set wolves population size at previous timestep
		
		# Calculate change in population size
		dE <- ((kp1*E) + (-pec*E*C) + (-pew*E*W)) * dt # Change in elk population size over dt
		dC <- ((-kd1*C) + (kp2*E*C) + (-kd2*C*W)) * dt # Change in coyotes population size over dt
		dW <- ((-kd3*W) + (kp3*E*W)) * dt # Change in wolves population size over dt
		
		# Calculate total population size
		E.simu[i] <- E + dE # Current elk population size at ith timestep
		C.simu[i] <- C + dC # Current coyotes population size at ith timestep
		W.simu[i] <- W + dW # Current wolves population size at ith timestep
		
		} # End of inner for loop
	
	# Store max/min population size for oscillation with stable asymptotic dynamics
	max_E.starset[j] <- max(E.simu[(length(E.simu)/2):(length(E.simu))])
	min_E.starset[j] <- min(E.simu[(length(E.simu)/2):(length(E.simu))])
	max_W.starset[j] <- max(W.simu[(length(W.simu)/2):(length(W.simu))])
	min_W.starset[j] <- min(W.simu[(length(W.simu)/2):(length(W.simu))])
	
} # End of outer foor loop
```

```{r}
# Plot max value of elk population against predation rate of wolves on elk (pew_set)
plot(x = pew_set, # Add pew_set vector to x axis
     y = max_E.starset, # Add max elk population size to y axis
     type = "l", # Set type to "Lines"
     lwd = 2, # Set line width to 2
     col = Ecol, # Set line color
     las = 1, # Change orientation of tick marks to be horizontal
     main = "Bifurcation for State Variable, E", # Set main title
     sub = "Time Series Simulation Adapted from Johnson et al. (2015) for Wolf Reintroduction in 1995",
     xlab = "Predation Rate of Wolves on Elk, pew", # Set x axis label
     ylab = "Number of elk, E*") # Set y axis label
# Plot min value of elk population against predation rate of wolves on elk (pew_set)
lines(x = pew_set, # Add pew_set vector to x axiss
      y = min_E.starset, # Add min elk population size to y axis
      lwd = 2, # Set line width to 2
      col = "#1d3557") # Specify line color
legend("topright",  # Set legend position
       legend = c("Max", "Min"), # Set item labels
       lwd = 2, # Set line width to 2
       lty = c(1, 1), # Set item line types
       col = c(Ecol, "#1d3557")) # Set item colors

# Plot max value of elk population against predation rate of wolves on elk (pew_set)
plot(x = pew_set, # Add pew_set vector to x axis
     y = max_W.starset, # Add max elk population size to y axis
     type = "l", # Set type to "Lines"
     lwd = 2, # Set line width to 2
     col = Wcol, # Set line color
     las = 1, # Change orientation of tick marks to be horizontal
     main = "Bifurcation for State Variable, E", # Set main title
     sub = "Time Series Simulation Adapted from Johnson et al. (2015) for Wolf Reintroduction in 1995",
     xlab = "Predation Rate of Wolves on Elk, pew", # Set x axis label
     ylab = "Number of wolves, W*") # Set y axis label
# Plot min value of elk population against predation rate of wolves on elk (pew_set)
lines(x = pew_set, # Add pew_set vector to x axiss
      y = min_W.starset, # Add min elk population size to y axis
      lwd = 2, # Set line width to 2
      col = "#a7c957") # Specify line color
legend("topright",  # Set legend position
       legend = c("Max", "Min"), # Set item labels
       lwd = 2, # Set line width to 2
       lty = c(1, 1), # Set item line types
       col = c(Wcol, "#a7c957")) # Set item colors
```

### Growth Rate of Elk

```{r}
# Create x axis vector of range of `kp1` values to bifurcate 
kp1_set <- seq(from = 0, to = 25, by = 0.05) 

# Create empty holding vectors for state variables to store min/max population sizes 
min_E.starset <- NaN*pew_set # NaN vector of vector length pew_set to store min elk population size
max_E.starset <- NaN*pew_set # NaN vector of vector length pew_set to store max elk population size
min_W.starset <- NaN*pew_set # NaN vector of vector length pew_set to store min wolf population size
max_W.starset <- NaN*pew_set # NaN vector of vector length pew_set to store max wolf population size
```

```{r}
## Calculate equilibrium values for each value of kp1 in `kp1_set` ##
for(j in 1:length(kp1_set)){ # Iterate over number of items given by length of kp1_set; start outer for loop

	# kp1 <- kp1_set[j] # Assign jth element of kp1_set to kp1
	kp1 <- kp1_set[j]
	
	## Simulate model for jth element of kp1_set ##
	
	# Set initial condition
	E.simu <- NaN*tset # Create NaN holding vector for elk population
	E.simu[1] <- 1 # Set initial elk population size 
	C.simu <- NaN*tset # Create NaN holding vector for coyotes population
	C.simu[1] <- 1 # Set initial coyote population size 
	W.simu <- NaN*tset # Create NaN holding vector for wolves population
	W.simu[1] <- 1 # Set initial wolves population size 
	
	# Initiate for loop, start in second slot due to initial condition; start inner for loop
	for(i in 2:length(tset)){
	  
	  dt <- tset[i]-tset[i-1] # Change in time; timestep
		
		# Create temporary variable from previous timestep
		E <- E.simu[i-1] # Set elk population size at previous timestep
		C <- C.simu[i-1] # Set coyotes population size at previous timestep
		W <- W.simu[i-1] # Set wolves population size at previous timestep
		
		# Calculate change in population size
		dE <- ((kp1*E) + (-pec*E*C) + (-pew*E*W)) * dt # Change in elk population size over dt
		dC <- ((-kd1*C) + (kp2*E*C) + (-kd2*C*W)) * dt # Change in coyotes population size over dt
		dW <- ((-kd3*W) + (kp3*E*W)) * dt # Change in wolves population size over dt
		
		# Calculate total population size
		E.simu[i] <- E + dE # Current elk population size at ith timestep
		C.simu[i] <- C + dC # Current coyotes population size at ith timestep
		W.simu[i] <- W + dW # Current wolves population size at ith timestep
		
		} # End of inner for loop
	
	# Store max/min population size for oscillation with stable asymptotic dynamics
	max_E.starset[j] <- max(E.simu[(length(E.simu)/2):(length(E.simu))])
	min_E.starset[j] <- min(E.simu[(length(E.simu)/2):(length(E.simu))])
	max_W.starset[j] <- max(W.simu[(length(W.simu)/2):(length(W.simu))])
	min_W.starset[j] <- min(W.simu[(length(W.simu)/2):(length(W.simu))])
	
} # End of outer for loop
```

```{r}
# Plot max value of elk population against growth rate of elk (kp1_set)
plot(x = kp1_set, # Add pew_set vector to x axis
     y = max_E.starset, # Add max elk population size to y axis
     type = "l", # Set type to "Lines"
     lwd = 2, # Set line width to 2
     col = Ecol, # Set line color
     las = 1, # Change orientation of tick marks to be horizontal
     main = "Bifurcation for State Variable, E", # Set main title
     sub = "Time Series Simulation Adapted from Johnson et al. (2015) for Wolf Reintroduction in 1995",
     xlab = "Growth Rate of Elk, kp1", # Set x axis label
     ylab = "Number of elk, E*") # Set y axis label
# Plot min value of elk population against growth rate of elk (kp1_set)
lines(x = kp1_set, # Add pew_set vector to x axiss
      y = min_E.starset, # Add min elk population size to y axis
      lwd = 2, # Set line width to 2
      col = "#1d3557") # Specify line color
legend("topright",  # Set legend position
       legend = c("Max", "Min"), # Set item labels
       lwd = 2, # Set line width to 2
       lty = c(1, 1), # Set item line types
       col = c(Ecol, "#1d3557")) # Set item colors

# Plot max value of elk population against growth rate of elk (kp1_set)
plot(x = kp1_set, # Add pew_set vector to x axis
     y = max_W.starset, # Add max elk population size to y axis
     type = "l", # Set type to "Lines"
     lwd = 2, # Set line width to 2
     col = Wcol, # Set line color
     las = 1, # Change orientation of tick marks to be horizontal
     main = "Bifurcation for State Variable, E", # Set main title
     sub = "Time Series Simulation Adapted from Johnson et al. (2015) for Wolf Reintroduction in 1995",
     xlab = "Growth Rate of Elk, kp1", # Set x axis label
     ylab = "Number of wolves, W*") # Set y axis label
# Plot min value of elk population against growth rate of elk (kp1_set)
lines(x = kp1_set, # Add pew_set vector to x axiss
      y = min_W.starset, # Add min elk population size to y axis
      lwd = 2, # Set line width to 2
      col = "#a7c957") # Specify line color
legend("topleft",  # Set legend position
       legend = c("Max", "Min"), # Set item labels
       lwd = 2, # Set line width to 2
       lty = c(1, 1), # Set item line types
       col = c(Wcol, "#a7c957")) # Set item colors
```

## Phase Diagram
```{r}
plot(x = W.simu, # Add kp1_set vector to x axis
     y = E.simu, # Add equilibrium elk population size to y axis
     type = "l", # Set type to "Lines"
     lwd = 2, # Set line width to 2
     col = 'aquamarine3', # Set line color
     las = 1, # Change orientation of tick marks to be horizontal
     xlim = c(min(c(E.simu, W.simu), na.rm = T),
             max(c(E.simu, W.simu), na.rm = T)),
     ylim = c(min(c(E.simu, W.simu), na.rm = T),
             max(c(E.simu, W.simu), na.rm = T)),
     xlab = "Elk Population", # Set x axis label
     ylab = "Wolf Population", # Set y axis label
     main = "Phase Diagram of Wolf Population vs Elk Population") # Set main title
```